---
title: "Untitled"
author: "Leo"
date: "2024-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following are exercises based on a toy example for showcasing ctsem's basic
capabilities by Charles Driver (https://cdriver.netlify.app/post/dtbivariate/).
(The code is simplified and adjusted to an exercise format with ctsemOMX.)
## Setup: Install & load required packages
```{r}
#install.packages("ctsem")
library(ctsem)
library(ctsemOMX)
```


## Simulate some data (ignore this)
```{r}
set.seed(1)

gm <- ctModel(LAMBDA=matrix(c(
  1,0,
  .8,0,
  0,1),ncol=2),
  MANIFESTVAR = diag(.5,3),
  DIFFUSION=matrix(c(.8,.5,0,.2),2,2),
  DRIFT=c(
    -.5,0,
    -.3,-.2),
  T0MEANS = c(20,-10),
  TRAITVAR=matrix(c(1,.7,0,.2),2,2),
  Tpoints=10,manifestNames = c('Stress','Stress2','Quality'))

df <-data.frame(ctGenerate(ctmodelobj = gm,n.subjects = 100,burnin = 2))

```

# Excercises
"For this example, we have collected data from workers starting a new job. 
The data include measures of stress, and quality of the employees output. 
We are interested in a) how stress and quality change from the beginning of the 
job, b) whether there is any evidence for one process causing the other." 
(see https://cdriver.netlify.app/post/dtbivariate/ for the full example)

## Excercise 0
Look at the data `df` and bring it in the correct format for the ctfit() function
(remember to use quotation for the column names)
```{r}

wideexample <- ctLongToWide(datalong =df , id ="id" ,
  time = "time" , manifestNames = c('Stress','Quality'))


df_wide <- ctIntervalise(datawide = wideexample, Tpoints = 10, n.manifest = 2,
                         manifestNames = c("Stress", "Quality"))

head(df_wide)
```

## Exercise 1
Set up a continuous SEM model using the `ctModel()` function that models the
bivariate (ignoring `Stress2` for now) in the data frame `df_wide`.
```{r}
example1model <- ctModel(n.latent = 2, n.manifest = 2, Tpoints = 10, 
                         manifestNames = c("Stress", "Quality"),
                         latentNames = c("Stress", "Quality"),
                         DRIFT = c('a_Stress','a_Stress_Quality',
                             'a_Quality_Stress','a_Quality'), # default
                         LAMBDA = diag(2))
```

## Exercise 2
Fit the model to the data using the ctfit() function.
```{r}
example1fit <- ctFit(datawide = df_wide, ctmodelobj = example1model)
```

## Exercise 3
Plot the resulting fit for the auto and cross-effects
```{r}
plot(example1fit, wait = FALSE)
```

## Exercise 4
Look at the fit summary (`summary(example1fit)`) and the plots. Interpret the model 
parameters and its fit in your own words. Do Stress and Quality influence each other?
How does this change as a function of the time interval considered? Which variable
is more stable?
```{r}
summary(example1fit)
```
### Cross-effects
Cross effects are interpreted analogous to regression effects:
An one unit increase of Stress leads to a 0.23 unit decrease in Quality.
A one unit increase of Quality leads to an 0.19 unit increase in Stress.

### Auto-effects
Auto effects are best interpreted in their converted form (by plotting) as 
solutions to the stochastic differential model equation as a function of the 
time interval.

However in general for the diagonal drift coefficients one can say:
The closer to 0 the drift parameter the more stable the process of the respective 
construct. The more negative a parameter gets the less stable the process. (This
can be understood when looking at the solution to the stochastic differential
equation)

Here we see Quality as a more stable construct across different
time intervals. While the auto-regressive effect of Stress decreases the longer
the respective time interval.


# Excercise 5 (optional)
Extend the model to account for unit heterogeneity by modelling intersubjective 
variability (random intercept) of manifest variables. 
Hint: simply add the `TRAITVAR = "auto"` argument to the model. Use the plot
function to graphically compare.
```{r}
traitmodel <- ctModel(n.latent = 2, n.manifest = 2, Tpoints = 10, 
                         manifestNames = c("Stress", "Quality"),
                         latentNames = c("Stress", "Quality"),
                         DRIFT = c('a_Stress','a_Stress_Quality',
                             'a_Quality_Stress','a_Quality'),
                         LAMBDA = diag(2),
                         TRAITVAR = "auto")
traitfit <- ctFit(datawide = df_wide, ctmodelobj = traitmodel)

plot(traitfit, wait = FALSE)
plot(example1fit, wait = FALSE)
```