---
title: "CM6 Hands on ctsem modelling - Excercise"
author: "Leonard Kosanke & Valentin Kriegmair"
date: "18-01-2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following are exercises based on a toy example for showcasing ctsem's
capabilities by Charles Driver (https://cdriver.netlify.app/post/dtbivariate/).
(The code is simplified and adjusted to an exercise format with ctsemOMX.)

## Setup: Install & load required packages
```{r}
#install.packages("ctsem")
library(ctsem)
library(ctsemOMX)
```

## Simulate some data (ignore this)
```{r}
set.seed(1)

# simulate data with 10 time points:
gm <- ctModel(LAMBDA=matrix(c(
  1,0,
  .8,0,
  0,1),ncol=2),
  MANIFESTVAR = diag(.5,3),
  DIFFUSION=matrix(c(.8,.5,0,.2),2,2),
  DRIFT=c(
    -.5,0,
    -.3,-.2),
  T0MEANS = c(20,-10),
  TRAITVAR=matrix(c(1,.7,0,.2),2,2),
  Tpoints=10,manifestNames = c('Stress','Stress2','Quality'))

df <-data.frame(ctGenerate(ctmodelobj = gm,n.subjects = 100,burnin = 2))
```

# Excercises
"For this example, we have collected data from workers starting a new job. 
The data include measures of stress, and quality of the employees output. 
We are interested in a) how stress and quality change from the beginning of the 
job, b) whether there is any evidence for one process causing the other." 
(see https://cdriver.netlify.app/post/dtbivariate/ for the full example)

## Excercise 0
Look at the data `df` and bring it in the correct format for the ctfit() function.

Here we just provide the solution feel free to just skip if you don't want to bother
with data wrangling right now.
```{r}
# from long format to wide format with absolute time values
df_wide0 <- ctLongToWide(datalong =df , id ="id" ,
  time = "time" , manifestNames = c('Stress', "Quality"))


# from absolute time points to time intervals (dT1 = T1 - T0 etc. ...)
df_wide <- ctIntervalise(datawide = df_wide0, Tpoints = 10, n.manifest = 2,
                         manifestNames = c("Stress", "Quality"))


# note in this toy example every subject has the same measurement time at each 
# wave for faster computations. 
# In ct modelling we reflect individually varying time points and
# intervals and using this even model more complex dynamics like oscillations.

head(df_wide)
```

## Exercise 1
Set up a continuous SEM model using the `ctModel()` function that models the
bivariate latent process (ignoring `Stress2` for now) in the data frame `df_wide`.
```{r}


```

## Exercise 2
Fit the model to the data using the `ctfit()` function.
```{r}

```

## Exercise 3
Plot the resulting fit for the auto and cross-effects (put `wait = F`)
```{r}

```

## Exercise 4
Look at the fit summary (`summary(example1fit)`) and the plots. Interpret the model 
parameters and its fit in your own words. Do Stress and Quality influence each other?
How does this change as a function of the time interval change? Which variable
is more stable?
```{r}

```
### Cross-effects:



### Auto-effects:


# Excercise 5 (optional)
Extend the model to account for unit heterogeneity by modelling intersubjective 
variability (random intercept) of manifest variables. 
Hint: simply add the `TRAITVAR = "auto"` argument to the model. Use the plot
function to graphically compare the model to the previous model. How does your
interpretation change?
```{r}


```


# Excercise 6 (optional)
Extend the `LAMBDA` matrix to include the second Stress indicator `Stress2`.
(Note matrices are filled rowise in ctsem)

first run this to extend the data set:
```{r}
# from long format to wide format with absolute time values
df_wide1 <- ctLongToWide(datalong =df , id ="id" ,
  time = "time" , manifestNames = c('Stress', "Stress2", "Quality"))


# from absolute time points to time intervals (dT1 = T1 - T0 etc. ...)
df_wide2 <- ctIntervalise(datawide = df_wide1, Tpoints = 10, n.manifest = 3,
                         manifestNames = c("Stress", "Stress2", "Quality"))


# note in this toy example every subject has the same measurement time at each 
# wave for faster computations. 
# In ct modelling we reflect individually varying time points and
# intervals and using this even model more complex dynamics like oscillations.

head(df_wide2)
```

Model with additional indicator `Stress2`:
```{r}


```

Fit the model:
```{r}
factorfit <- ctFit(datawide = df_wide2, ctmodelobj = factormodel)

summary(factorfit)
```
