---
title: "Hands On Continuous Time Modeling With ctsem"
author: "Leonard Kosanke & Valentin Kriegmair"
format: 
  revealjs:
    theme: "serif"
    slide-number: true
    incremental: true
    transition: "none" 

---

## Outline
* Recap
* Introduction to ctsem
* Walk-through
* Excercise
* Solution

## Recap I
What are the assumptions about time in the general cross lagged panel model (GCLPM)?

1. time prgoresses in discrete steps

2. equally spaced measurement occasions

3. subjects are measured at the same time points

## Recap II
What are limitations of treating time as a discrete variable when modeling longitudinal data?

* biased parameter estimates due to:

  * unequally spaced measurement waves in the panel data

  * measurement of individual subjects at each wave spread out over time

* uncomparable parameter estimates across studies

## Recap II
Multivariate stochastic differential equation:
\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left({\mathbf{A}} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}

## Recap II {visibility="uncounted"}
Multivariate stochastic \color{red}{differential} equation:
\begin{gather*}
\color{red}{\mathrm{d}} \boldsymbol{\eta}_i(\color{red}{t})=\left(\mathbf{A} \boldsymbol{\eta}_i(\color{red}{t})+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \color{red}{\mathrm{d} t}+\mathbf{G} \color{red}{\mathrm{d}} \mathbf{W}_i(\color{red}{t})
\end{gather*}
modelling change in latent variables $\boldsymbol{\eta}_i(t)$
 as a function of time over infinitesimal small time steps $\color{red}{\mathrm{d} t}$

## Recap II {visibility="uncounted"}
Multivariate stochastic differential equation:
\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\color{red}{\mathbf{A}} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}
modelling change as a function of time over infinitesimal small time steps $\mathrm{d} t$

$\color{red}{\mathbf{A}}$: drift matrix with auto effects on the diagonal and cross effects on the off-diagonals ($\mathbf{A} \in \mathbb{R}^{v \times v}$).


## Recap II {.smaller visibility="uncounted"}
Multivariate stochastic differential equation:
\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\mathbf{A} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}
modelling change as a function of time over infinitesimal small time steps $\mathrm{d} t$

$\mathbf{A}$: drift matrix with auto effects on the diagonal and cross effects on the off-diagonals ($\mathbf{A} \in \mathbb{R}^{v \times v}$).

$\mathbf{z_i}$: time independent predictors

$\boldsymbol{\chi}_i(t)$: time dependent predictors

$\mathbf{W}_i(t)$: stochastic error term (Wiener process)

$\mathbf{G}$: diffusion matrix

## Recap II {.smaller}
Multivariate stochastic differential equation:
\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\mathbf{A} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}
measurement part:
\begin{gather*}
\mathbf{y}(t)=\boldsymbol{\Lambda} \boldsymbol{\eta}(t)+{\boldsymbol{\tau}}+\boldsymbol{\varepsilon}(t)
\end{gather*}
\begin{gather*}
\boldsymbol{\varepsilon}(t) \sim \mathrm{N}(\mathbf{0}, \mathbf{Θ})
\end{gather*}

$\boldsymbol{\Lambda}$: factor loadings

$\boldsymbol{\tau}$: intercepts

$\boldsymbol{\varepsilon}(t)$: measurement error

## {background-image="figures/figure3-1.png" background-size="contain" background-repeat="no-repeat"}
## {background-image="figures/figure3-2.png" background-size="contain" background-repeat="no-repeat"}

## Introducing `ctsem` & `ctsemOMX` {.smaller}
* R-packages `ctsem` & `ctsemOMX` to handle the math
* `ctsem` interfaces with Stan
* `ctsemOMX` interfaces with OpenMx
* today we will use `ctsemOMX`
```{r}
# install.packages("ctsem")
# install.packages("ctsemOMX")
library(ctsem) # loads ctsem
library(ctsemOMX) # loads ctsem and ctsemOMX
```

## Data Structure {.smaller auto-animate="true"}
::: {.fragmet}
`ctsemOMX` uses wide data with time intervals specified in columns:
```{r}
#| output: false
set.seed(1)

# simulate data with 10 time points:
gm <- ctModel(LAMBDA=matrix(c(
  1,0,
  .8,0,
  0,1),ncol=2),
  MANIFESTVAR = diag(.5,3),
  DIFFUSION=matrix(c(.8,.5,0,.2),2,2),
  DRIFT=c(
    -.5,0,
    -.3,-.2),
  T0MEANS = c(20,-10),
  TRAITVAR=matrix(c(1,.7,0,.2),2,2),
  Tpoints=10,manifestNames = c('Stress','Stress2','Quality'))

df <-data.frame(ctGenerate(ctmodelobj = gm,n.subjects = 100,burnin = 2))
df_wide0 <- ctLongToWide(datalong =df , id ="id" ,
  time = "time" , manifestNames = c('Stress', "Quality"))

df_wide <- ctIntervalise(datawide = df_wide0,
            Tpoints = 10, n.manifest = 2,
            manifestNames = c("Stress", "Quality"))

dataframewide<- as.data.frame(df_wide)
```
```{r}
library(tidyverse)
dataframewide <- dataframewide %>% select( Quality_T0, Quality_T1,Quality_T2, dT1:dT3)
head(dataframewide)
```

(time points can also be interindividually varying)
:::

::: {.fragmet}
Function to convert to wideformat:
```r
df_wide <- ctLongToWide(datalong = df_long, id = "id",
  time = "time", manifestNames = c("Quality", ...))
```
:::

::: {.fragmet}
Function to create time intervals:
```r
df_wide <- ctIntervalise(datawide = wideexample, Tpoints = 3, n
  manifest = 2, manifestNames = c("Quality", ...))
```
:::

# Next: Model set-up and interpretation walkthrough

## {background-image="figures/figure2-1.png" background-size="contain" background-repeat="no-repeat"}